---
# title: |
#  | \LARGE{\textbf{Junkyu Park}}
#  | \tiny{\ }
#  | \small{1603-580 The East Mall, Etobicoke, ON M9B 4A7}
#  | \small{647-676-4374}
#  | \small{joonjunkyu.park@mail.utoronto.ca}
#  | \small{https://www.linkedin.com/in/asdfzxcvjkl1/}
#  | \small{https://joon3216.github.io/}
header-includes:
 - \usepackage{multicol}
 - \usepackage{xcolor}
 - \definecolor{linkblue}{RGB}{30, 107, 184}
 - \renewcommand{\baselinestretch}{.85}
 # - \newcommand{\code}[1]{\texttt{#1}}
output: 
  pdf_document:
    toc_depth: 1
fontsize: 11pt
geometry: "left=.75in,right=.75in,top=.7in,bottom=.7in"
---

<!-- Last updated: 2021-01-22 -->


```{r beginmulticols21, echo = F, results = 'asis'}
cat("
\\begin{multicols}{2}
\\raggedcolumns
")
```

\begin{minipage}{\linewidth}
\LARGE{\textbf{Junkyu Park}}
\end{minipage}

\begin{minipage}{\linewidth}
\begingroup
\fontsize{10pt}{10pt}
\texttt{joonjunkyu.park@alum.utoronto.ca}

\texttt{\textcolor{linkblue}{\url{https://www.linkedin.com/in/asdfzxcvjkl1/}}}

\texttt{\textcolor{linkblue}{\url{https://joon3216.github.io/}}}
\endgroup
\end{minipage}


```{r endmulticols1, echo = F, results = 'asis'}
cat("\\end{multicols}")
```

# Skills

* Data analysis
    + data manipulation: `dplyr`, `tidyr`, `pandas`
    + data visualization: `ggplot2`, `matplotlib.pyplot`, `seaborn`
    + regression analysis: R functions and packages, `statsmodels.api`
    + statistical computations: manually written functions, `numpy`, `scipy`
    + comprehension: able to find references and read documentations of external packages
* Technical
    + Markdown/R Markdown: able to form documentations
    + Git and git repositories: git commands, GitLab, GitHub
    + Some knowledge of HTML and CSS: currently maintaining \textcolor{linkblue}{\href{https://joon3216.github.io/}{\texttt{joon3216.github.io}}}
    + SQL: able to write complex queries and subqueries
    + Python and R
        + able to utilize external packages for modeling, computations, and visualizations
        + able to develop a simple package and upload to archives and git repositories
* Language: English, Korean

```{r beginmulticols2, echo = F, results = 'asis', eval = F}
cat("
\\begin{multicols}{2}
\\raggedcolumns
")
```

<!-- \begin{minipage}{.99\linewidth} -->
<!--     \textbf{Data analysis} -->
<!--         \begin{itemize} -->
<!--             \item Data manipulation -->
<!--                 \begin{itemize} -->
<!--                     \item \texttt{pandas}, \texttt{patsy} -->
<!--                     \item \texttt{dplyr}, \texttt{tidyr} -->
<!--                     \item SQL queries -->
<!--                 \end{itemize} -->
<!--             \item Data visualization -->
<!--                 \begin{itemize} -->
<!--                     \item \texttt{matplotlib.pyplot}, \texttt{seaborn} -->
<!--                     \item \texttt{ggplot2} -->
<!--                 \end{itemize} -->
<!--             \item Regression analysis -->
<!--                 \begin{itemize} -->
<!--                     \item \texttt{statsmodels.api} -->
<!--                 \end{itemize} -->
<!--             \item Statistical computations -->
<!--                 \begin{itemize} -->
<!--                     \item \texttt{numpy}, \texttt{scipy} -->
<!--                 \end{itemize} -->
<!--         \end{itemize} -->
<!--     \smallskip -->
<!--     \textbf{Languages} -->
<!--         \begin{itemize} -->
<!--             \item English -->
<!--             \item Korean -->
<!--         \end{itemize} -->
<!-- \end{minipage} -->

<!-- \begin{minipage}{.99\linewidth} -->
<!--     \textbf{Technical} -->
<!--         \begin{itemize} -->
<!--             \item Markdown/R Markdown -->
<!--                 \begin{itemize} -->
<!--                     \item able to integrate dynamic badges supported by Travis CI and Codecov.io -->
<!--                 \end{itemize} -->
<!--             \item Some knowledge of HTML and CSS -->
<!--                 \begin{itemize} -->
<!--                     \item currently maintaining \textcolor{linkblue}{\href{https://joon3216.github.io/}{joon3216.github.io}} -->
<!--                 \end{itemize} -->
<!--             \item PostgreSQL -->
<!--                 \begin{itemize} -->
<!--                     \item able to use keywords such as \texttt{unnest}, \texttt{array}, and \texttt{information\_schema} -->
<!--                     \item able to join tables and write subqueries -->
<!--                 \end{itemize} -->
<!--             \item Python -->
<!--                 \begin{itemize} -->
<!--                     \item currently maintaining \textcolor{linkblue}{\href{https://github.com/joon3216/statspark}{\texttt{statspark}}} and \textcolor{linkblue}{\href{https://github.com/joon3216/sudsoln}{\texttt{sudsoln}}} packages -->
<!--                 \end{itemize} -->
<!--             \item R -->
<!--             \begin{itemize} -->
<!--                 \item currently maintaining \textcolor{linkblue}{\href{https://github.com/joon3216/funpark}{\texttt{funpark}}} package -->
<!--             \end{itemize} -->
<!--         \end{itemize} -->
<!-- \end{minipage} -->

```{r endmulticols, echo = F, results = 'asis', eval = F}
cat("\\end{multicols}")
```


# Projects

**Package development** (work): `estime2` package in Python

* built a prototype of the program that will replace the current system used by Demography Division that fixes negative end-of-period populations in various regional-level population tables
* incorporated various methods to fix a negative end-of-period population and evaluation metrics for each of those methods


**Package development** (work): `sprcr` package in R

* included the following functionalities to the package:
    + plotting kernel density estimates and empirical cdfs based on different colour and panel groups
    + smoothing out data points based on constraints and various configurations
    + performing k-fold cross-validation in the smoothing process

<!-- **Modeling** (personal): credit card fraud detection -->

<!-- * built statistical models and wrote an \textcolor{linkblue}{\href{https://joon3216.github.io/projects/2019/creditcardfraud/analysis}{analysis}} about the data and the detection of fraudulent credit card uses -->
<!-- * built a model that is expected to achieve an accuracy of 97.105% and a true positive rate of 87.204% for every 6-hour worth of transaction data with the current model and the training set -->



# Experience

<!-- **Volunteer note-taker** \hspace*{\fill} Sep 2016 - Apr 2019   -->
<!-- University of Toronto Accessibility Services \hspace*{\fill} 2 years 8 months -->

<!-- * volunteered as a note-taker for students with disabilities -->

<!-- \smallskip -->

```{r duration_calc, echo = F, warning = F, message = F}
library(lubridate)
library(zoo)
month_writer <- function(testing, writing) {
  
    ifelse(
        testing == 0,
        writing,
        ifelse(
            testing == 1, '1 mo', 
            paste(as.character(testing), 'mos')
        )
    )
}
to_ym <- function(day_start, day_end = today()) {
  
    result <- (as.yearmon(day_end) - as.yearmon(day_start)) * 12
    result <- round(result)
    if (result < 12) {
        month_writer(result, '1 mo')
    } else {
        y <- result %/% 12
        m <- result - 12 * y
        paste0(
            paste(as.character(y), ifelse(y == 1, 'yr', 'yrs')),
            ifelse(m == 0, '', ' '),
            month_writer(m, '')
        )
    }
}
```

**Methodologist** (MA-01) \hspace*{\fill} Jan 2020 - Present $\cdot$ `r to_ym('2020-01-06')`  
Statistics Canada \hspace*{\fill} Ottawa, ON

* wrote documentations on various topics and ideas in statistics and computations
* conducted analysis and modeling requested by clients and/or senior methodologists
* developed, and maintaining, packages wrtten in R and Python




<!-- # Awards -->

<!-- * Dean's List Scholar \hspace*{\fill} 2013 - 2014, 2016 - 2019 -->
<!-- * Louis Savlov Scholarship \hspace*{\fill} Oct 2017 -->
<!-- * Awards for Excellence in Army Training Test \hspace*{\fill} May 2016 -->
<!-- * Math Honour Roll \hspace*{\fill} Jun 2014 -->
<!-- * University of Toronto Entrance Scholarship \hspace*{\fill} Oct 2013 -->


# Education

**H.B.Sc.: statistics specialist and math minor** \hspace*{\fill} 2019  
University of Toronto  


<!-- **Certifications**   -->
<!-- Society of Actuaries (SOA) -->

<!-- * Exams: P (passed), FM (passed), SRM (accredited) -->
<!-- * All VEEs obtained: Economics, Applied Statistics, Corporate Finance -->





