---
# title: |
#  | \LARGE{\textbf{Junkyu Park}}
#  | \tiny{\ }
#  | \small{1603-580 The East Mall, Etobicoke, ON M9B 4A7}
#  | \small{647-676-4374}
#  | \small{joonjunkyu.park@mail.utoronto.ca}
#  | \small{https://www.linkedin.com/in/asdfzxcvjkl1/}
#  | \small{https://joon3216.github.io/}
header-includes:
 - \usepackage{multicol}
 - \usepackage{xcolor}
 - \definecolor{linkblue}{RGB}{30, 107, 184}
 - \renewcommand{\baselinestretch}{.85}
 # - \newcommand{\code}[1]{\texttt{#1}}
output: 
  pdf_document:
    toc_depth: 1
fontsize: 11pt
geometry: margin = .75in
---

<!-- Last updated: 2019-06-25 -->

<!-- \begingroup -->
<!-- \fontsize{10pt}{12pt} -->
<!-- \begin{verbatim} -->
<!-- 1603-580 The East Mall, Etobicoke, ON M9B 4A7   -->
<!-- 647-676-4374   -->
<!-- joonjunkyu.park@mail.utoronto.ca   -->
<!-- https://www.linkedin.com/in/asdfzxcvjkl1/   -->
<!-- https://joon3216.github.io/   -->
<!-- \end{verbatim} -->
<!-- \endgroup -->

```{r beginmulticols21, echo = F, results = 'asis'}
cat("
\\begin{multicols}{2}
\\raggedcolumns
")
```

\begin{minipage}{\linewidth}
\LARGE{\textbf{Junkyu Park}}
\end{minipage}

\begin{minipage}{\linewidth}
\begingroup
\fontsize{10pt}{10pt}
\texttt{1603-580 The East Mall, Etobicoke, ON M9B 4A7}

\texttt{647-676-4374}

\texttt{joonjunkyu.park@mail.utoronto.ca}

\texttt{\textcolor{linkblue}{\url{https://www.linkedin.com/in/asdfzxcvjkl1/}}}

\texttt{\textcolor{linkblue}{\url{https://joon3216.github.io/}}}
\endgroup
\end{minipage}


```{r endmulticols1, echo = F, results = 'asis'}
cat("\\end{multicols}")
```


# Education

**H.B.Sc.: statistics specialist and math minor** (GPA: 3.83) \hspace*{\fill} Jun 2019  
University of Toronto  

\smallskip

**Certifications**  
Society of Actuaries (SOA)

* Exams: P (passed), FM (passed), SRM (accredited)
* All VEEs obtained: Economics, Applied Statistics, Corporate Finance


# Experience

**Volunteer note-taker** \hspace*{\fill} Sep 2016 - Apr 2019  
University of Toronto Accessibility Services \hspace*{\fill} 2 years 8 months

* volunteered as a note-taker for students with disabilities

\smallskip

**Assistant Squad Leader** (Sergeant) \hspace*{\fill} Oct 2014 - Jul 2016  
The 15^th^ Infantry Division, Republic of Korea Army \hspace*{\fill} 1 year 9 months

* deployed the second time to the General Outpost from Jun 2016 until discharged
* assumed the position of Assistant Squad Leader as of Dec 2015 until discharged
* deployed to the 38^th^ Regiment's General Outpost from Sep 2015 to Feb 2016


# Skills

```{r beginmulticols2, echo = F, results = 'asis'}
cat("
\\begin{multicols}{2}
\\raggedcolumns
")
```

\begin{minipage}{.99\linewidth}
\textbf{Data analysis}
\begin{itemize}
  \item Data manipulation
  \begin{itemize}
    \item aggregating, reshaping, and tidying
  \end{itemize}
  \item Data visualization
  \item Regression analysis
  \item Statistical computations
  \item Quantitative methods
\end{itemize}

\bigskip

\textbf{Languages}: English, Korean

\bigskip

\textbf{Technical}: Git, Google Analytics
\end{minipage}

\begin{minipage}{.99\linewidth}
\textbf{Programming}
\begin{itemize}
  \item Some knowledge of HTML and CSS
  \begin{itemize}
    \item currently maintaining \textcolor{linkblue}{\href{https://joon3216.github.io/}{joon3216.github.io}}
  \end{itemize}
  \item PostgreSQL (moderately familiar)
    \begin{itemize}
      \item able to use keywords such as \texttt{unnest}, \texttt{array}, and \texttt{information\_schema}
      \item able to join tables and write subqueries
    \end{itemize}
  \item Python (competent)
  \item R (proficient)
    \begin{itemize}
      \item currently maintaining \textcolor{linkblue}{\href{https://github.com/joon3216/funpark}{\texttt{funpark} package}}
    \end{itemize}
\end{itemize}
\end{minipage}

```{r endmulticols, echo = F, results = 'asis'}
cat("\\end{multicols}")
```


# Awards

* Dean's List Scholar \hspace*{\fill} 2013 - 2014, 2016 - 2019
* Louis Savlov Scholarship \hspace*{\fill} Oct 2017
* Awards for Excellence in Army Training Test \hspace*{\fill} May 2016
* Math Honour Roll \hspace*{\fill} Jun 2014
* University of Toronto Entrance Scholarship \hspace*{\fill} Oct 2013











