---
title: "Introduction to sudsoln"
author: Junkyu Park
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    includes:
      in_header: 
        - ../../../style/all_ga_script.html
        - ../../../style/all_navbar_head.html
        - ../../../style/all_orange_jp_03_lvl.html
      before_body:
        - ../../../style/all_navbar_body_03_lvl.html
      after_body:
        - ../../../style/all_footer_03_lvl.html
    toc: TRUE
    self_contained: FALSE
---

```{r setup, echo = F, results = 'hide'}
library(reticulate)
knitr::knit_engines$set(python = eng_python)
use_virtualenv('r-reticulate')
py_available(T)
```

This vignette focuses on how you can initialize `Sudoku` puzzle and use methods to solve them. Any other details, such as classes like `Candidate`, `Array`, `Appearance`, or other `Sudoku` methods such as `.__eq__()`, `.group()`, `.copy()` and so on, are omitted.


To install `sudsoln`, type:

```
pip install sudsoln
```

Note that `sudsoln` is only supported in Python 3.

In this document, the following version of `sudsoln` is imported with the following alias:

```{python import_sudsoln}
import sudsoln as ss
print(ss.__version__)
```

```{python under_the_hood1, echo = F}
ss_ver = ss.__version__
```

```{r under_the_hood2, echo = F}
ss_ver <- py$ss_ver
```

# Initializing a puzzle

There are two ways to initialize a puzzle: 

* use `Sudoku` class directly
* use `to_sudoku` function

The difference between the two is that `Sudoku` accepts a list of lists in its `array` argument, whereas `to_sudoku` function accepts a string in its `sudoku_str` argument and converts it into a list of lists. For example, if you have the following puzzle:

```{python eg1}
q1_array = [
    ['.', '.', '.', '.', '2', '.', '.', '.', '.'],
    ['8', '3', '.', '7', '1', '4', '.', '9', '6'],
    ['.', '6', '.', '9', '.', '5', '4', '.', '8'],
    ['.', '9', '.', '3', '.', '1', '.', '.', '4'],
    ['.', '1', '.', '4', '.', '2', '.', '.', '7'],
    ['.', '7', '5', '.', '.', '.', '2', '1', '.'],
    ['.', '.', '4', '.', '.', '.', '7', '.', '.'],
    ['.', '.', '.', '5', '.', '7', '.', '.', '.'],
    ['.', '.', '.', '1', '9', '6', '.', '.', '.']
]
```

then you have to use `Sudoku` class. If you have it in a form of string as follows:

```{python eg2}
q1_str = '....2....83.714.96.6.9.54.8.9.3.1..4.1.4.2..7.75...21...4...7.....5.7......196...'
```

then you should use `to_sudoku` function. `to_sudoku` is merely a wrapper function of `Sudoku`; inside `to_sudoku`, `q1_str` is converted to the same form as `q1_array`, and this will be plugged into `array` argument of `Sudoku` along other arguments.

Both require three arguments in total. Two common arguments are:

* `empty`: a string that denotes the emptiness (`'.'` by default)
* `elements`: a set of entries in an array that is NOT `empty`

Usually, if a size of puzzle is 9-by-9 (or 3^2^-by-3^2^), then `elements` is a set of integers from `1` to `9`. Unlike `empty`, `elements` has no default argument.

If `elements` is not specified in either `Sudoku` or `to_sudoku`, then it guesses `elements` from: 

* `array` of `Sudoku`, or; 
* `sudoku_str` of `to_sudoku`

depending on what you are using to initialize one. For example, both `q1_array` and `q1_str` will be successfully initialized without any specification of either `empty` or `elements`:

```{python eg3}
q1_case1 = ss.Sudoku(q1_array)
q1_case1
q1_case2 = ss.to_sudoku(q1_str)
q1_case2
print(q1_case1 == q1_case2)
```

The following `q4_str`, however, will not be initialized properly and will return `ValueError` with the following message:

```{python eg4}
q4_str = ss.questions.q4
print(q4_str)
```

```{python eg4_trial, eval = F}
q4 = ss.to_sudoku(q4_str)
```



```
ValueError: Length of the guessed elements is 8, not 9. Either make sure that: 
1. every element in the current array contains all of your intended elements at
least once, or; 2. specify elements explicitly, or; 3. there is exactly one
string, and only one, that denotes the emptiness in the array. For example, if
you try to solve a 9-by-9 sudoku whose answer form consists of integers from 1 
to 9, either make sure that every integer from 1 to 9 shows up in the current
array at least once, or explicitly specify elements = set([str(i) for i in
range(1, 10)]), or see if the array uses '.' and some other string, like ',' or 
' ', to denote the emptiness.
```

This is because `Sudoku` thinks any element in `array` that is NOT `empty` should be a member of `elements`. `9` does not exist in `q4_str`, so it causes an error that says the number of guessed elements is 8, not 9. The solution of this problem is to specify `elements` as follows:

```{python eg5}
q4_case1 = ss.to_sudoku(q4_str, elements = set([i for i in range(1, 10)]))
q4_case1
```

In the error message, it suggests to use `set([str(i) for i in range(1, 10)])`, but it is fine to use `set([i for i in range(1, 10)])`; `Sudoku` automatically converts every element in `elements` into `str` before creating an instance of `Sudoku` puzzle anyway.

If you don't like `'.'` as `empty` in your puzzle, you have to change it BEFORE initializing into a `Sudoku` puzzle. That is, the following is not acceptable:

```{python eg6}
print(q1_str) # recall q1_str
```

```{python eg6_1, eval = F}
q1_case3 = ss.to_sudoku(q1_str, empty = ' ')
```

```
KeyError: "empty = ' ' does not exist in the array. Either specify the correct
string denoting the emptiness in the array, or change the string denoting the
emptiness in the array by using sudsoln.change_empty(array, old, new)."
```

Instead, you have to do the following:

```{python eg7}
q1_str = q1_str.replace('.', ' ')
print(q1_str)
q1_case3 = ss.to_sudoku(q1_str, empty = ' ')
q1_case3
```

or the following in case of `q4_str`:

```{python eg8}
q4_str = q4_str.replace('.', '*')
print(q4_str)
q4_case2 = ss.to_sudoku(
    q4_str, 
    elements = set([i for i in range(1, 10)]), 
    empty = '*'
)
q4_case2
```

After successful initialization, you are ready to use methods to solve the puzzle.

Note: every element in `elements` as well as `empty` must have length 1 if it is converted into type `str`. For example, `10` or `14` cannot be an element in `elements`, or `'..'` cannot be `empty`.



# Solving a puzzle

As of version `r ss_ver`, there are 7 solving methods provided:

* `Sudoku.solve_globally()`
* `Sudoku.solve_locally()`
* `Sudoku.solve_by_pointing_pairs()`
* `Sudoku.solve_by_hidden_pairs()`
* `Sudoku.solve_logically()`
* `Sudoku.solve_forcefully()`
* `Sudoku.solve()`

## `.solve_globally()`

For each entry, this method looks at the `row`, `col`umn, and `submatrix` it belongs and see if there is only one possible candidate value at that entry. If true, plug the value into that entry, and repeat the process until no new mutation is made to `Sudoku`.

For example, take a look at `q1_case1`:

```{python sg_eg1}
q1_case1
```

See the entry at the 3^rd^ row and 5^th^ column. Notice that `3` is the only possible candidate value at that entry because:

* there are `4`, `5`, `6`, `8`, and `9` in the row it belongs
* there are `1` and `2` in the column it belongs
* there is `7` in the submatrix it belongs

And therefore, the only candidate value left at that entry is `3`, indicating it must be the value at that entry. `.solve_globally()` does this for every entry in the sudoku array, and goes back to the very first entry after it goes through the entire array. The method stops if the process does not mutate `Sudoku` puzzle anymore.

`q1_case1` is actually a very easy sudoku puzzle and only requires `.solve_globally()` to be solved entirely:

```{python sg_eg2}
from datetime import datetime as dt
start = dt.now(); q1_case1.solve_globally(); end = dt.now()
q1_case1
str(end - start)
```

`q4_case1`, however, is a harder puzzle. You can see that `.solve_globally()` doesn't do anything:

```{python sg_eg3}
q4_case1_cp = q4_case1.copy()
q4_case1.solve_globally()
q4_case1
print(q4_case1 == q4_case1_cp)
```

## `.solve_locally()`

This looks at candidate values more carefully. It doesn't look at the only candidate value remaining; rather, it tries to acquire the only existing candidate value within the local group it belongs, i.e. `row`, `col`, or `submatrix`. If there is only one number that is exclusively a candidate value in one particular entry within a local group, that entry must have that value in its place. Just like `.solve_globally()`, `.solve_locally()` runs until no further mutation is made in `Sudoku` puzzle.

It has to take the argument `by`, the name of a local group that it will work with. `by` must be either `row`, `col`, or `submatrix`

Here's an example:

```{python slocal_eg1}
q_small = ss.questions.q_small
print(q_small)
```

`q_small` is a 4-by-4 (or 2^2^-by-2^2^) sudoku puzzle:

```{python slocal_eg2}
q_small = ss.to_sudoku(q_small)
q_small
```

Take a look at the 2^nd^ row and candidate values in each entry of it. The entry at the 2^nd^ row and:

* the 1^st^ column has `3` and `4` as candidate values because:
    + `2` is in the same row
    + `1` is in the same column
    + `1` and `2` are in the same submatrix
* the 3^rd^ column has `1` and `4` as candidate values because:
    + `2` is in the same row
    + `3` is in the same column and submatrix
* the 4^th^ column has `1` as the only candidate value because:
    + `2` is in the same row
    + `4` is in the same column
    + `3` is in the same submatrix

Although the entry at the 2^nd^ row and the 1^st^ column has two candidate values (`3` and `4`), we can conclude that `3` must be the value at this entry because `3` is the only candidate value in the 2^nd^ row that exists at this entry only. 

`q2_case1` is an example of a puzzle that needs `.solve_locally()` to get to the answer form:

```{python slocal_eg3}
q2_case1 = ss.to_sudoku(ss.questions.q2)
q2_case2 = q2_case1.copy()

# Case 1: solve locally by row then column 
q2_case1
q2_case1.solve_locally(by = 'row')
q2_case1.solve_locally(by = 'col')
q2_case1
```

```{python slocal_eg4}
# Case 2: solve locally by submatrix
q2_case2
q2_case2.solve_locally(by = 'submatrix')
q2_case2
```

## `.solve_by_pointing_pairs()`

As of version 0.1.0, this only works with `by = 'submatrix'`. Working methods for `by = 'row'` and `by = 'col'` will be supported in version 0.1.1.

Inside a submatrix, if a single row or column exclusively contains one candidate number, then that candidate number should be eliminated from the same row or column that is not a part of the submatrix.

Here is an example:

```{python q_top95_4_eg1}
pzl = '4873...9....6..271126.9.3847.5...162...2..8..........9..1.769233..1..45.....53.18'
pzl = ss.to_sudoku(pzl, elements = set([i for i in range(1, 10)]))
pzl
```

See the candidate values in the fourth submatrix:

```{python qt4_eg2}
pzl.group(by = 'submatrix')[4] # 1-indexing for submatrix
```

See the entries with keys `(5, x)` (i.e. entries of the 6^th^ row in the sudoku puzzle that belong to the 4^th^ submatrix). Notice that these entries exclusively have `8` as a candidate value within this submatrix. This means any entry that:

1. belongs to the 6^th^ row, and
2. does NOT belong to the 4^th^ submatrix, and
3. has `8` as one of its candidate values

should eliminate `8` from one of its candidate values. That is, among those entries of the 6^th^ row:

```{python pzl_eg3}
pzl.group(by = 'row')[5] # 0-indexing for row and col
```

the entries `(5, 3)`, `(5, 4)`, and `(5, 5)` should get rid of `8` from one of its candidate values since they do not belong to the 4^th^ submatrix and has `8` as one of its candidate values.


## `.solve_by_hidden_pairs()`

Within a local group (`row`, `col`, or `submatrix`), if two distinct entries has two candidate values exclusively, then these two entries should only contain those two values in its set of candidate values.

For example, consider the following `Candidate` group:

```{python pzl_eg4}
ss.candidate.Candidate( # 6th row
    {
        (5, 0): {'6', '8', '2'},
        (5, 1): {'3', '6', '4', '1'},
        (5, 2): {'3', '4', '8', '2'},
        (5, 3): {'5', '7'},
        (5, 4): {'3', '4', '1', '6'},
        (5, 5): {'4', '1'},
        (5, 6): {'5', '7'},
        (5, 7): {'3', '4'},
    },
    elements = set([i for i in range(1, 10)])
)
```

See `(5, 0)` and `(5, 2)`. These two entries contain `2` and `8` exclusively, so the set of candidate values should be `{'2', '8'}` for both.


## `.solve_logically()`

This method applies:

* `.solve_globally()`
* `.solve_locally(by)`, where `by = 'row'`, `'col'`, and `'submatrix'`
* `.solve_by_pointing_pairs()`
* `.solve_by_hidden_pairs(by)`, where `by = 'row'`, `'col'`, and `'submatrix'`

It continues to apply these methods until either no mutation is made to `Sudoku`, or `Sudoku` transforms into the answer form.

## `.solve_forcefully()`

In this method, the following steps are repeated:

1. Apply `.solve_logically()` and get the most refined version of `Candidate` from `Sudoku`.
2. From the first entry (i.e. the entry with the smallest `row` and `col` in `(row, col)`) of `Sudoku` that is `empty`, randomly plug in one value from the candidate set.
3. Apply `.solve_logically()` after plugging in.
4. Repeat 2 and 3 until:
    + 4.1. `Sudoku` transforms into the answer form, OR
    + 4.2. at least one candidate set of `Sudoku` entries is `set()`, i.e. the empty set.

In case of 4.2., `Sudoku` transforms back to the initial state, the method adds `1` to the `trial` number and goes back to step 1. This process continues until either `trial == max_trial` or `Sudoku` transforms into the answer form.


## `.solve()`

This method first uses `.solve_logically()`, and then `.solve_forcefully()`.


# Performance

See [analysis](analysis.html) for details.



