---
title: "Introduction to sudsoln"
author: Junkyu Park
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    includes:
      in_header: 
        - ../../../style/all_ga_script.html
        - ../../../style/all_navbar_head.html
        - ../../../style/all_orange_jp_03_lvl.html
      before_body:
        - ../../../style/all_navbar_body_03_lvl.html
      after_body:
        - ../../../style/all_footer_03_lvl.html
    toc: TRUE
    self_contained: FALSE
---

```{r setup, echo = F, results = 'hide'}
library(reticulate)
knitr::knit_engines$set(python = eng_python)
use_virtualenv('r-reticulate')
py_available(T)
```

(Note: `sudsoln` is only supported in Python 3)

To install `sudsoln`, type:

```
pip install sudsoln
```

This vignette focuses on how you can initialize `Sudoku` puzzle and use methods to solve them. Any other details, such as classes like `Candidate`, `Array`, `Appearance`, or other `Sudoku` methods such as `.__eq__()`, `.group()`, `.copy()` and so on, are omitted.

In this document, the following version of `sudsoln` is imported with the following alias:

```{python import_sudsoln}
import sudsoln as ss
print(ss.__version__)
```

```{python under_the_hood1, echo = F}
ss_ver = ss.__version__
```

```{r under_the_hood2, echo = F}
ss_ver <- py$ss_ver
```

# Initializing a puzzle

There are two ways to initialize a puzzle: 

* use `Sudoku` class directly
* use `to_sudoku` function

The difference between the two is that `Sudoku` accepts a list of lists in `array` argument, whereas `to_sudoku` function accepts a string in `sudoku_str` argument. 

In fact, `to_sudoku` is merely a wrapper function around `Sudoku`; inside `to_sudoku`, `sudoku_str` is converted to a list of lists, and this is plugged into `array` argument of `Sudoku`.

Both `Sudoku` and `to_sudoku` require three arguments in total:

* For `Sudoku`, they are `array`, `empty`, and `elements`
* For `to_sudoku`, they are `sudoku_str`, `empty`, and `elements`

Details regarding these arguments are written in step 2.

## Step 0: check if you have a compatible size/length

`sudsoln` is designed to handle $n^2$-by-$n^2$ sudoku puzzles, where $n \geq 2$ is a number of rows/columns in a subarray (or `submatrix`, the term used in this package). You need to make sure that a puzzle has a size/length $n^4$, where $n \geq 2$ is a natural number. Compatible sizes/lengths include:

* $16 = 2^4$,
* $81 = 3^4$,
* $256 = 4^4$,
* $625 = 5^4$

and so forth.


## Step 1: try it without any additional arguments

You can first try initializing sudoku puzzle without specifying any additional arguments (i.e. `empty` and `elements`). Suppose you have the following puzzle:

```{python eg1}
q1_array = [
    ['.', '.', '.', '.', '2', '.', '.', '.', '.'],
    ['8', '3', '.', '7', '1', '4', '.', '9', '6'],
    ['.', '6', '.', '9', '.', '5', '4', '.', '8'],
    ['.', '9', '.', '3', '.', '1', '.', '.', '4'],
    ['.', '1', '.', '4', '.', '2', '.', '.', '7'],
    ['.', '7', '5', '.', '.', '.', '2', '1', '.'],
    ['.', '.', '4', '.', '.', '.', '7', '.', '.'],
    ['.', '.', '.', '5', '.', '7', '.', '.', '.'],
    ['.', '.', '.', '1', '9', '6', '.', '.', '.']
]
```

You may use `Sudoku` directly as follows:

```{python eg1_1}
q1_case1 = ss.Sudoku(q1_array)
q1_case1
```

If you have it in a form of string:

```{python eg2}
q1_str = '....2....83.714.96.6.9.54.8.9.3.1..4.1.4.2..7.75...21...4...7.....5.7......196...'
```

then you should use `to_sudoku` function as follows:

```{python eg2_1}
q1_case2 = ss.to_sudoku(q1_str)
q1_case2
```

Both initializations are successful without specifying `empty` and `elements`.


## Step 2: specify `elements`

As mentioned earlier, both `Sudoku` and `to_sudoku` require three arguments in total. Two common arguments are:

* `empty`: a string that denotes the emptiness (`'.'` by default)
* `elements`: a set of entries in `array`/`sudoku_str` that is NOT `empty`

Notes:

1. Unlike `empty`, `elements` has no default argument.
2. Usually, if a size of puzzle is 9-by-9 (or 3^2^-by-3^2^), then `elements` is a set of integers from `1` to `9`.
3. Every element in `elements` as well as `empty` must have length 1 if it is converted into type `str`. For example, `10` or `14` cannot be an element of `elements`, or `'..'` cannot be `empty`.

If `elements` is not specified, then it *guesses* `elements` from `array` or `sudoku_str` depending on what you are using to initialize a puzzle. Specifically, `Sudoku` thinks any element in `array` that is NOT `empty` should be a member of `elements`. The reason for successful initializations of `q1_array` and `q1_str` in step 1 is because it has 9 distinct values --- integers from 1 to 9 --- in the `array` that is NOT `empty` (i.e. `'.'`).

That is, any puzzle that does NOT have the "correct" number of distinct values will not be initialized in step 1. For example, `q4_str`, will not be initialized properly and will return `ValueError` with the following message:

```{python eg4}
q4_str = ss.questions.q4
print(q4_str)
```

```{python eg4_trial, eval = F}
q4_case1 = ss.to_sudoku(q4_str)
```

```
ValueError: Length of the guessed elements is 8, not 9. Either make sure that: 
1. every element in the current array contains all of your intended elements at
least once, or; 2. specify elements explicitly, or; 3. there is exactly one
string, and only one, that denotes the emptiness in the array. For example, if
you try to solve a 9-by-9 sudoku whose answer form consists of integers from 1 
to 9, either make sure that every integer from 1 to 9 shows up in the current
array at least once, or explicitly specify elements = set([str(i) for i in
range(1, 10)]), or see if the array uses '.' and some other string, like ',' or 
' ', to denote the emptiness.
```

Initialization is not successful because `9` does not exist in `q4_str`, so it causes an error that says the number of guessed elements is 8, not 9.

The solution of this problem is to specify `elements` as follows:

```{python eg5}
q4_case1 = ss.to_sudoku(
    q4_str, 
    elements = set([i for i in range(1, 10)])
)
q4_case1
```

Notice that I used `set([i for i in range(1, 10)])` although the message suggests to use `set([str(i) for i in range(1, 10)])`. You can go with either of these as `Sudoku` automatically converts every element in `elements` into `str` before creating an instance of `Sudoku` puzzle anyway.

## Step 3: specify the correct `empty` string

If you don't like `'.'` as `empty` in your puzzle, you have to change it BEFORE initializing into a `Sudoku` puzzle (I may provide the `empty` setter of `Sudoku` in future updates). That is, the following won't work:

```{python eg6}
print(q1_str) # recall q1_str
```

```{python eg6_1, eval = F}
q1_case3 = ss.to_sudoku(q1_str, empty = ' ')
```

```
KeyError: "empty = ' ' does not exist in the array. Either specify the correct
string denoting the emptiness in the array, or change the string denoting the
emptiness in the array by using sudsoln.change_empty(array, old, new)."
```

You see this error message because `q1_str` uses a period (`'.'`) to denote the emptiness. Instead, you have to do the following:

```{python eg7}
q1_str = q1_str.replace('.', ' ')
print(q1_str)
q1_case3 = ss.to_sudoku(q1_str, empty = ' ')
q1_case3
```

You may do the following in case of `q4_str`:

```{python eg8}
q4_str = q4_str.replace('.', '*')
print(q4_str)
q4_case2 = ss.to_sudoku(
    q4_str, 
    elements = set([i for i in range(1, 10)]), 
    empty = '*'
)
q4_case2
```



## Step 4: fix typos

Suppose you are heading towards the subway station commuting to your workplace. Along the way, you grab a newspaper, read two or three articles, criticize journalists on their political orientations, and then get to the end of the newspaper. You see today's 3^2^-by-3^2^ sudoku puzzle is right there, remembering there is a Python package called `sudsoln` you downloaded before, and you decide to give it a try.

You open your laptop/smartphone/tablet, open a Python shell (on [Pydroid](https://play.google.com/store/apps/details?id=ru.iiec.pydroid3&hl=en_CA)/[Termux](https://play.google.com/store/apps/details?id=com.termux&hl=en_CA)/etc in case of smartphone or tablet), and then type in the following:

```{python step4_eg1}
puzzle = '..3.,..2....416..56.8.......2.9.4.6..6..8..7..3.2.7.9.......6.42..841....5....9..'
```

You then `import sudsoln as ss`, and try initialize `Sudoku` using `to_sudoku` as follows:

```{python step4_eg2, eval = F}
puzzle = ss.to_sudoku(puzzle)
```

But you get the error message:

```
ValueError: Length of the guessed elements is 10, not 9. Either make sure that: 
1. every element in the current array contains all of your intended elements at
least once, or; 2. specify elements explicitly, or; 3. there is exactly one
string, and only one, that denotes the emptiness in the array. For example, if
you try to solve a 9-by-9 sudoku whose answer form consists of integers from 1 
to 9, either make sure that every integer from 1 to 9 shows up in the current
array at least once, or explicitly specify elements = set([str(i) for i in
range(1, 10)]), or see if the array uses '.' and some other string, like ',' or 
' ', to denote the emptiness.
```

Ah, you remember this error message. The writer of this package said that you have to specify `elements` in this case:

```{python step4_eg3, eval = F}
puzzle = ss.to_sudoku(puzzle, elements = set([i for i in range(1, 10)]))
```

But this time, you are getting a different error message:

```
ValueError: There exists an element in array that is not a member of 
elements: {','}
```

What does this mean? This means that `puzzle` has a string that is neither `empty` nor an element of `elements`. Specifically, the error message says that `','` is in your `puzzle` which is not supposed to be an element of `elements`, but not `empty` either.

You scroll up the shell to the point where you defined `puzzle`, and realize that the 5^th^ string of `puzzle` is `','` and not `'.'`. You must have made a mistake while typing the puzzle and have not realized it. After all, a comma and a period is right next to each other:

```{python step4_eg4}
puzzle[4]
```

You also see that the first error message said the "[l]ength of the guessed elements is **10**, not 9." The guessed elements was not 8 but 10 after all.

Now you know the mistake, you fix the typo and then initialize it:

```{python step_eg5}
puzzle = puzzle.replace(',', '.')
puzzle = ss.to_sudoku(puzzle)
puzzle
```

After successful initialization, you are ready to use methods to solve the puzzle.



# Solving a puzzle

As of version `r ss_ver`, there are 7 solving methods provided:

* `Sudoku.solve_globally()`
* `Sudoku.solve_locally()`
* `Sudoku.solve_by_pointing_pairs()`
* `Sudoku.solve_by_hidden_pairs()`
* `Sudoku.solve_logically()`
* `Sudoku.solve_forcefully()`
* `Sudoku.solve()`

## `.solve_globally()`

This method accepts no argument, and returns `None`.

For each entry, this method looks at the `row`, `col`umn, and `submatrix` it belongs and see if there is only one possible candidate value at that entry. If true, plug the value into that entry, and repeat the process until no new mutation is made to `Sudoku`.

For example, take a look at `q1_case1`:

```{python sg_eg1}
q1_case1
```

See the entry at the 3^rd^ row and 5^th^ column. Notice that `3` is the only possible candidate value at that entry because:

* there are `4`, `5`, `6`, `8`, and `9` in the row it belongs
* there are `1` and `2` in the column it belongs
* there is `7` in the submatrix it belongs

And therefore, the only candidate value left at that entry is `3`, indicating it must be the value at that entry. `.solve_globally()` does this for every entry in the sudoku array, and goes back to the very first entry after it goes through the entire array. The method stops if the process does not mutate `Sudoku` puzzle anymore.

`q1_case1` is actually a very easy sudoku puzzle and only requires `.solve_globally()` to be solved entirely:

```{python sg_eg2}
from datetime import datetime as dt
start = dt.now(); q1_case1.solve_globally(); end = dt.now()
q1_case1
str(end - start)
```

`q4_case1`, however, is a harder puzzle. You can see that `.solve_globally()` doesn't do anything:

```{python sg_eg3}
q4_case1_cp = q4_case1.copy()
q4_case1.solve_globally()
q4_case1
print(q4_case1 == q4_case1_cp)
```

## `.solve_locally()`

This method has `by` argument, the name of a local group that it will work with. It must be specified, and must be one of either `row`, `col`, or `submatrix`. It returns `None`.

This looks at candidate values more carefully. It doesn't look at the only candidate value remaining; rather, it tries to acquire the only existing candidate value within the local group it belongs, i.e. `row`, `col`, or `submatrix`. If there is only one number that is exclusively a candidate value in one particular entry within a local group, that entry must have that value in its place. Just like `.solve_globally()`, `.solve_locally()` runs until no further mutation is made in `Sudoku` puzzle.


Here's an example:

```{python slocal_eg1}
q_small = ss.questions.q_small
print(q_small)
```

`q_small` is a 4-by-4 (or 2^2^-by-2^2^) sudoku puzzle:

```{python slocal_eg2}
q_small = ss.to_sudoku(q_small)
q_small
```

Take a look at the 2^nd^ row and candidate values in each entry of it:

```{python slocal_eg2_1}
q_small.group(by = 'row')[1] # 0-indexing for row and col
```

The entry at the 2^nd^ row and:

* the 1^st^ column (i.e. entry `(1, 0)`) has `3` and `4` as candidate values because:
    + `2` is in the same row
    + `1` is in the same column
    + `1` and `2` are in the same submatrix
* the 3^rd^ column (i.e. entry `(1, 2)`) has `1` and `4` as candidate values because:
    + `2` is in the same row
    + `3` is in the same column and submatrix
* the 4^th^ column (i.e. entry `(1, 3)`) has `1` as the only candidate value because:
    + `2` is in the same row
    + `4` is in the same column
    + `3` is in the same submatrix

Although the entry at the 2^nd^ row and the 1^st^ column has two candidate values (`3` and `4`), we can conclude that `3` must be the value at this entry because `3` is the only candidate value in the 2^nd^ row that exists at this entry only. 

`puzzle` in step 4 above is an example of a puzzle that needs `.solve_locally()` to get to the answer form:

```{python slocal_eg3}
q2_case1 = puzzle
q2_case2 = q2_case1.copy()

# Case 1: solve locally by row then column 
q2_case1
q2_case1.solve_locally(by = 'row')
q2_case1.solve_locally(by = 'col')
q2_case1
```

```{python slocal_eg4}
# Case 2: solve locally by submatrix
q2_case2
q2_case2.solve_locally(by = 'submatrix')
q2_case2
```

## `.solve_by_pointing_pairs()`

(As of version 0.1.0, this only works with `by = 'submatrix'`. Working methods for `by = 'row'` and `by = 'col'` will be supported in version 0.1.1)

This method takes:

* `by` argument, `submatrix` by default, and;
* `start` argument, `None` by default, which is an initial `Candidate` that the method will start to reduce.

It returns a `Candidate` class reduced down to the furthest level achievable by the method itself. If `start` is specified, then this method will not only mutate `Sudoku`, but also mutate `start` AND return the reduced version of `start`.

Inside a submatrix, if a single row or column exclusively contains one candidate number, then that candidate number should be eliminated from the same row or column that is not a part of the submatrix.

Here is an example:

```{python q_top95_4_eg1}
pzl = '4873...9....6..271126.9.3847.5...162...2..8..........9..1.769233..1..45.....53.18'
pzl = ss.to_sudoku(pzl, elements = set([i for i in range(1, 10)]))
pzl
```

See the candidate values in the fourth submatrix:

```{python qt4_eg2}
pzl.group(by = 'submatrix')[4] # 1-indexing for submatrix
```

See the entries with keys `(5, x)` (i.e. entries of the 6^th^ row in the sudoku puzzle that belong to the 4^th^ submatrix). Notice that `8` exclusively exists in the candidate value sets of entries `(5, x)` within this submatrix. This means any entry that:

1. belongs to the 6^th^ row, and
2. does NOT belong to the 4^th^ submatrix, and
3. has `8` as one of its candidate values

should eliminate `8` from one of its candidate values. That is, among those entries of the 6^th^ row:

```{python pzl_eg3}
pzl.group(by = 'row')[5] # Again, 0-indexing for row and col
```

the entries `(5, 3)`, `(5, 4)`, and `(5, 5)` should get rid of `8` from one of its candidate values since they do not belong to the 4^th^ submatrix and has `8` as one of its candidate values.


## `.solve_by_hidden_pairs()`

This method takes:

* `by` argument, `submatrix` by default, and;
* `start` argument, `None` by default, which is an initial `Candidate` that the method will start to reduce.

This method works with all possible `by`s, which are `row`, `col`, and `submatrix`. And it returns a `Candidate` class reduced down to the furthest level achievable by the method itself. If `start` is specified, then this method will not only mutate `Sudoku`, but also mutate `start` AND return the reduced version of `start`.

Within a local group `by` (`== row`, `col`, or `submatrix`), if two distinct entries have two candidate values exclusively, then these two entries should only contain those two values in its set of candidate values.

For example, consider the following `Candidate` group:

```{python pzl_eg4}
candids = ss.candidate.Candidate( # 6th row
    {
        (5, 0): {'6', '8', '2'},
        (5, 1): {'3', '6', '4', '1'},
        (5, 2): {'3', '4', '8', '2'},
        (5, 3): {'5', '7'},
        (5, 4): {'3', '4', '1', '6'},
        (5, 5): {'4', '1'},
        (5, 6): {'5', '7'},
        (5, 7): {'3', '4'},
    },
    elements = set([i for i in range(1, 10)])
)
```

See `(5, 0)` and `(5, 2)`. These two entries contain `2` and `8` exclusively in this group, so the set of candidate values should be `{'2', '8'}` for both.

```{python pzl_eg5, echo = F, eval = F}
etm = ss.candidate.Candidate(
    {}, 
    elements = set([i for i in range(1, 10)])
)
candids_appear = candids.appearances(['col', 'submatrix'])
candids_appear.sieve(condition = ['contains', 2], deep = True)
candids.refine(
    entries_to_mutate = etm,
    appearances = candids_appear,
    condition = ['contains', 2],
    deep = True
)
candids
```

## `.solve_logically()`

This method takes no argument. It returns either `None` or `Candidate`, depending on the difficulty level of `Sudoku` puzzle. If `Sudoku` is solved only by `Sudoku.solve_globally()` and `Sudoku.solve_locally()` (i.e. easy puzzle), then it returns `None`, and returns `Candidate` otherwise.

This method applies:

* `.solve_globally()`
* `.solve_locally(by)`, where `by = 'row'`, `'col'`, and `'submatrix'`
* `.solve_by_pointing_pairs()`
* `.solve_by_hidden_pairs(by)`, where `by = 'row'`, `'col'`, and `'submatrix'`

It continues to apply these methods until either no mutation is made to `Sudoku`, or `Sudoku` transforms into the answer form.

## `.solve_forcefully()`

This method takes:

* `max_trial`, `200` by default, which is the maximum number of attempts to be tried
* `quietly`, `False` by default, which prints the message while solving forcefully
* `seed` (`None` by default) which is a natural number to put into `random.seed()`. If `None`, then `random.seed()` is not set.

It returns `int`, the number of attempts tried.

In this method, the following steps are repeated:

1. Apply `.solve_logically()` and get the most refined version of `Candidate` from `Sudoku`.
2. From the first entry (i.e. the entry with the smallest `row` and `col` in `(row, col)`) of `Sudoku` that is `empty`, randomly plug in one value from the candidate set.
3. Apply `.solve_logically()` after plugging in.
4. Repeat 2 and 3 until:
    + 4.1. `Sudoku` transforms into the answer form, OR
    + 4.2. at least one candidate set of `Sudoku` entries is `set()`, i.e. the empty set.

In case of 4.2., `Sudoku` transforms back to the initial state, the method adds `1` to the `trial` number and goes back to step 1. This process continues until either `trial == max_trial` (`200` by default) or `Sudoku` transforms into the answer form.


## `.solve()`

This method has `max_trial`, `quietly`, and `seed` arguments, which are the same arguments as `Sudoku.solve_forcefully()`. It returns a tuple of `(str, int)`, where the first item is the string representation of `datetime.timedelta` (e.g. `'0:00:03.140492'`), and the second item is the return value of `Sudoku.solve_forcefully()`. That is, the second item will be `0` if a puzzle is solved by `Sudoku.solve_logically()` only.

This method first uses `.solve_logically()`, and then `.solve_forcefully()`.


# Performance

See [analysis](analysis.html) for details.



